---
title: "class09"
author: "Rocio Barahona"
date: "May 1, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r}

wisc.df <- read.csv("https://bioboot.github.io/bimm143_S18/class-material/WisconsinCancer.csv")

# Convert the features of the data: wisc.data
wisc.data <- as.matrix(wisc.df[,3:ncol(wisc.df)])

# Set the row names of wisc.data
row.names(wisc.data) <- wisc.df$id

wisc.data <- wisc.data[,-31]

```


How many observations have a malignant diagnosis?
```{r}
table(wisc.df$diagnosis)
```

How many observations are in this dataset? 569 patients
```{r}
nrow(wisc.df)
```

 
```{r}

# Create diagnosis vector by completing the missing code

diagnosis <- as.numeric(wisc.df$diagnosis == "M")
table(diagnosis)

```

```{r}
# Convert the features of the data: wisc.data

wisc.data <- as.matrix(wisc.df[,3:ncol(wisc.df)])
rownames(wisc.data) <- wisc.df$id
head(wisc.data)


```

How many variables/features in the data are suffixed with _mean?
```{r}
grep("mean", colnames(wisc.df), value = TRUE)
```





#Performing PCA

```{r}

# Check column means and standard deviations
colMeans(wisc.data)

apply(wisc.data,2,sd)

```


```{r}

# Perform PCA on wisc.data by completing the following code
wisc.pr <- prcomp(wisc.data[,1:30], scale = TRUE)

# Look at summary of results
summary(wisc.pr)

```

Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?
44%


Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?


Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?




```{r}
attributes(wisc.pr)
```




```{r}
biplot(wisc.pr)
```

Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?
not easy to understand because it's all squished.

```{r}

# Scatter plot observations by components 1 and 2
#plot(wisc.pr, col = 1:2 , 
     #xlab = "PC1", ylab = "PC2")
# Repeat for components 1 and 2
plot(wisc.pr$x[, c(1, 2)], col = (diagnosis + 1), 
     pch = as.vector(wisc.df$diagnosis),
     xlab = "PC1", ylab = "PC2")

```



Q8. Repeat the same for principal components 1 and 3. What do you notice about these plots?

```{r}

# Repeat for components 1 and 3
plot(wisc.pr$x[, c(1, 3)], col = (diagnosis + 1), xlab = "PC1", ylab = "PC3")

```

Varience Explained: Scree Plot
```{r}

# Calculate varience of each component
pr.var <- wisc.pr$sdev^2
head(pr.var)

# Proportion of varience
pve <- pr.var / sum(pr.var)

# Plot varience explained for each principal component
plot(wisc.pr$sdev^2, xlab = "Principal Component", 
     ylab = "Cumulative Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")

```



```{r}

# Alternative scree plot of the same data, note data driven y-axis
barplot(pve, ylab = "Percent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )

```


Selecting number of clusters

```{r}

#?hclust
#?dist

# Scale the wisc.data data: data.scaled
data.scaled <- scale(wisc.data)


# Calculate the the distance matrix needed for hclust
data.dist <- dist(data.scaled)


wisc.hclust <- hclust(data.dist)


wisc.hclust.clusters <- cutree(wisc.hclust, h = 20)

# plot our hclust model tree
plot(wisc.hclust)

  
```


```{r}
table(wisc.hclust.clusters, diagnosis)
```


K-means


```{r}

# Scale  your wisc.data and peform kmeans
data.scaled <- scale(wisc.data)

#wisc.km <- kmeans(data.scaled, 2, nstart= 20)

```


```{r}
# compare the cluster membership of the k-means model to the actual diagnoses
table(wisc.km$cluster, diagnosis)

```

Q13. How well does k-means separate the two diagnoses? How does it compare to your hclust results?

```{r}
# Compare the cluster membership of the k-means model to your hierarchical clustering model from above
table(wisc.km$cluster, wisc.hclust.clusters)
```


```{r}
wisc.pr.hclust <- hclust(wisc.km, method = "complete" )

plot(wisc.pr.hclust)

wisc.pr.hclust.clusters <- cutree(wisc.pr.hclust, k=4)

table(wisc.pr.hclust.clusters, diagnosis)
```












